---
title: "synData9_LSM25"
format: html
editor: visual
---

```{r}
#| echo: false
#| output: false

# Housekeeping
source("https://raw.githubusercontent.com/jeffzifanwu/rProject/main/procedures/beginScript.R")
options(digits = 4)

# load library
library(dplyr)
library(dataHelper)
library(ggplot2)
source("https://raw.githubusercontent.com/jeffzifanwu/rProject/main/procedures/setGgplotThemes.R")
library(tidyr)
library(cowplot)
```

## Setup

```{r}
# Parameters
alpha <- 0.5    # weight on consumption
beta  <- 1   # weight on leisure
gamma <- 0.2    # disutility of labour
w     <- 1     # hourly wage rate
te    <- 16     # total time endowment
me    <- -2    # monetary endowment
u     <- 3

# Functions
committedLabour <- function(w, me) -me/w
consumption     <- function(H, w) w * H
leisure         <- function(H, HC, te) te - H - HC
totalLabour     <- function(H, HC) H + HC

# Utility function
utility <- function(H, w, me, te) {
  HC <- committedLabour(w, me)
  C <- consumption(H, w)
  L <- leisure(H, HC, te)
  T <- totalLabour(H, HC)

  if (C < 0 || L <= 0 || T < 0) return(-Inf)  # infeasible solution
  # debug print
  # cat("C, L, T, HC = ", C, L, T, HC, "\n")
  # cat("alpha * log(C)  = ", alpha * log(C) , "\n")
  # cat("beta * log(L) = ", beta * log(L), "\n")
  # cat("gamma * log(T) = ", gamma * log(T), "\n")
  
  # return
  alpha * log(C) + beta * log(L) + gamma * log(T)
}

# check scale
utility(4, w, me, te)


```

### Constraint space

```{r}
HC = committedLabour(w, me)
df.budget = data.frame(leisure = seq(1, te-HC, by = 0.1)) %>%
  mutate(committed_labour= HC) %>%
  mutate(uncommitted_labour = te - leisure - committed_labour) %>%
  mutate(consumption = w * uncommitted_labour) %>%
  mutate(total_labour = uncommitted_labour + committed_labour)

varLevels = c(
  "consumption",
  "committed_labour",
  "uncommitted_labour",
  "total_labour"
  )

df.plot.budget <- df.budget %>%
  pivot_longer(cols = c(uncommitted_labour, consumption, committed_labour, total_labour), names_to = "variable", values_to = "value") %>%
  mutate(variable = factor(variable, levels = varLevels))

# Plot
ggp = ggplot(df.plot.budget, aes(x = leisure, y = value)) +
  geom_line(color = "seagreen4") +
  geom_vline(xintercept = te, linetype = "dashed", color = "seagreen4") +
  geom_vline(xintercept = te - HC, linetype = "dotted", color = "seagreen4") +
  facet_wrap(~ variable, scales = "free_y", ncol = 2) +
  labs(x = "Leisure", y = NULL)

print(ggp)

```

### Utility maximisation problem

We test our utility maximisation problem

```{r}
# ump <- optimize(function(H) utility(H, w, me, te), interval = c(0.01, te - HC - 0.01), maximum = TRUE)
uncompensatedLabour <- function(w, me, te) {
  # Compute committed labour
  HC <- committedLabour(w, me)
  
  # Maximise utility subject to time constraint
  ump <- optimize(
    f = function(H) utility(H, w, me, te),
    interval = c(0.01, te - HC - 0.01),
    maximum = TRUE
  )
  
  # Get optimal H
  H <- ump$maximum
  
  maxU <- ump$objective 
  # print(maxU)
  
  # Return both HC and H
  return(H)
}

# Extract optimal values
HC <- committedLabour(w, me)
ump.H <- uncompensatedLabour(w, me, te)
ump.C <- consumption(ump.H, w)
ump.L <- leisure(ump.H, HC, te)
ump.T <- totalLabour(ump.H, HC)
ump.U <- utility(ump.H, w, me, te)

# Print results
cat("  Time endowment:", te, "\n")
cat("  HC: ", HC, "\n")
cat("  ump.H:", ump.H, "\n")
cat("  ump.U:", ump.U, "\n")
```

### Time resource minimiation

```{r}
# minTimeEndowment function: utilityTarget is only included for the syntax in nloptr. The current calculation doesn't need it.
minTimeEndowment <- function(x, w, me, utilityTarget) {
  HC <- committedLabour(w, me)
  if (x[1] <= 0) return(1e6)  # Penalise infeasible
  if (x[2] <= 0) return(1e6)
  x[1] + x[2] + HC
}

# check: it should return te
x = c(ump.H, ump.L)
minTimeEndowment(x, w, me)

# Utility function - form 2 - takes L and H inputs
tc.utility <- function(x, w, me) {
  
  # x[1] is H, x[2] is L
  
  HC <- committedLabour(w, me)
  C <- consumption(x[1], w)
  T <- totalLabour(x[1], HC)

  if (C <= 0 || x[2] <= 0 || T <= 0 || x[1] <= 0) return(-Inf)  # infeasible solution
  # alpha * log(C) + beta * log(x[2]) + gamma * log(T)
  utility(x[1], w, me, T + x[2])
}

# check if tc.utility returns ump.U
tc.utility(x, w, me)

# Constraint: utility >= utilityTarget
tc.utilityConstraint <- function(x, w, me, utilityTarget) {
  utilityTarget - tc.utility(x, w, me)
}

# check: it should return 0
tc.utilityConstraint(x, w, me, ump.U)

# Optimise: time compensated labour supply
library(nloptr)
compensatedLabour <- function(w, me, utilityTarget) {
  
  # minimisation
  rmp <- nloptr(
    x0 = c(1,1), # initial guess i.e., starting value
    w = w, me = me,  utilityTarget = utilityTarget, # other inputs
    eval_f = minTimeEndowment,
    eval_g_ineq = tc.utilityConstraint,
    lb = c(0.01, 0.01),
    ub = c(Inf, Inf),
    opts = list("algorithm" = "NLOPT_LN_COBYLA", "xtol_rel" = 1e-6)
  )
  
  # Extract optimal values
  rmp.sol <- rmp$solution
  rmp.obj <- rmp$objective
  
  # check objective value
  # cat("Objective value:", rmp.obj, "\n")
  
  # Return compensated H
  return(rmp.sol)
}

rmp.sol = compensatedLabour(w, me, ump.U)
rmp.H   = rmp.sol[1]
rmp.L   = rmp.sol[2]
teMin = minTimeEndowment(rmp.sol, w, me)
rmp.U  = tc.utility(rmp.sol, w, me)

# Results
cat("  Target utility =", ump.U, "\n")
cat("  Achieved utility =", rmp.U, "\n")
cat("  rmp.H:",  ump.H, "\n")
cat("  rmp.L:",  ump.L, "\n")
cat("  rmp.te:", teMin, "\n")

```

### Monetary resource minimisation problem

```{r}
# minMonetaryEndowment function: utilityTarget is only included for the syntax in nloptr. The current calculation doesn't need it.
minMonetaryEndowment <- function(x, w, te, utilityTarget) {
  # x[1] = H, x[2] = L
  if (x[1] <= 0) return(1e6)  # Penalise infeasible
  if (x[2] <= 0) return(1e6)
  w * (x[1] + x[2] - te)
}

# check: it should return me
x = c(ump.H, ump.L)
minMonetaryEndowment(x, w, te)

# Utility function - form 3 - takes L, H and te inputs
mc.utility <- function(x, w, te) {
  
  # x[1] is H, x[2] is L
  
  HC <- te - x[1] - x[2]
  C <- consumption(x[1], w)
  T <- totalLabour(x[1], HC)
  
  # cat("HC: ", HC, "\n")
  # cat("H: ", x[1], "\n")
  
  if (C <= 0 || x[2] <= 0 || T <= 0 || x[1] <= 0) return(-Inf)  # infeasible solution
  # utility(H, w, me, te)
  utility(x[1], w, (-1) * w * HC, te)
}

# check if mc.utility returns ump.U
mc.utility(x, w, te)

# Constraint: utility >= utilityTarget
mc.utilityConstraint <- function(x, w, te, utilityTarget) {
  utilityTarget - mc.utility(x, w, te)
}

# check: it should return 0
mc.utilityConstraint(x, w, te, ump.U)

# Optimise: monetary compensated labour supply
library(nloptr)
mcLabour <- function(w, te, utilityTarget) {
  
  # minimisation
  rmp <- nloptr(
    x0 = c(1,1), # initial guess i.e., starting value
    w = w, te = te,  utilityTarget = utilityTarget, # other inputs
    eval_f = minMonetaryEndowment,
    eval_g_ineq = mc.utilityConstraint,
    lb = c(0.01, 0.01),
    ub = c(Inf, Inf),
    opts = list("algorithm" = "NLOPT_LN_COBYLA", "xtol_rel" = 1e-6)
  )
  
  # Extract optimal values
  rmp.sol <- rmp$solution
  rmp.obj <- rmp$objective
  
  # check objective value
  # cat("Objective value:", rmp.obj, "\n")
  
  # Return compensated H
  return(rmp.sol)
}

rmp.sol = mcLabour(w, te, ump.U)
rmp.H   = rmp.sol[1]
rmp.L   = rmp.sol[2]
meMin = minMonetaryEndowment(rmp.sol, w, te)
rmp.U  = mc.utility(rmp.sol, w, te)

# Results
cat("  Target utility =", ump.U, "\n")
cat("  Achieved utility =", rmp.U, "\n")
cat("  rmp.H:",  ump.H, "\n")
cat("  rmp.L:",  ump.L, "\n")
cat("  rmp.me:", meMin, "\n")

```

# ------------------

# Simulate curves: Cobb-Douglas

# -----------------

```{r}

# simulate wage rates, time endowment, monetary endowment
teSeq <- seq(12, 20, length.out = 20)
meSeq <- seq(-8, -0.5, length.out = 20)
wSeq  <- seq(0.4, 1.6, length.out = 20)

# time endowment
df.te <- do.call(rbind, lapply(teSeq, function(x) {
  # for each value in the sequence, compute:
  H <- uncompensatedLabour(w, me, x)[1]
  HC<- committedLabour(w, me)
  T <- totalLabour(H, HC)
  U <- utility(H, w, me, x)
  # create a dataframe for the current values
  data.frame(
    wage_rate = w, 
    monetary_endowment = me, 
    time_endowment = x, 
    committed_labour = HC,
    uncommitted_labour = H,
    total_labour = T,
    utility = U
    )
}))

# monetary endowment
df.me <- do.call(rbind, lapply(meSeq, function(x) {
  # for each value in the sequence, compute:
  H <- uncompensatedLabour(w, x, te)[1]
  HC<- committedLabour(w, x)
  T <- totalLabour(H, HC)
  U <- utility(H, w, x, te)
  # create a dataframe for the current values
  data.frame(
    wage_rate = w, 
    monetary_endowment = x, 
    time_endowment = te, 
    committed_labour = HC,
    uncommitted_labour = H,
    total_labour = T,
    utility = U
    )
}))

# wage rate
df.w <- do.call(rbind, lapply(wSeq, function(x) {
  # for each value in the sequence, compute:
  H <- uncompensatedLabour(x, me, te)[1]
  HC<- committedLabour(x, me)
  T <- totalLabour(H, HC)
  U <- utility(H, x, me, te)
  # create a dataframe for the current values
  data.frame(
    wage_rate = x, 
    monetary_endowment = me, 
    time_endowment = te, 
    committed_labour = HC,
    uncommitted_labour = H,
    total_labour = T,
    utility = U
    )
}))

# Monetary-compensated time endowment
df.mcTe = do.call(rbind, lapply(teSeq, function(x){
  # for each value in the sequence, compute:
  sol <- mcLabour(w, x, u)
  H   <- sol[1]
  meMin <- minMonetaryEndowment(sol, w, x)
  HC  <- committedLabour(w, meMin)
  T   <- totalLabour(H, HC)
  # create a dataframe for the current values
  data.frame(
    wage_rate = w, 
    time_endowment = x, 
    utility_level = u, 
    committed_labour = HC,
    uncommitted_labour = H,
    total_labour = T,
    monetary_resource = meMin
    )
}))

# Monetary-compensated wage rate
df.mcWr = do.call(rbind, lapply(wSeq, function(x){
  # for each value in the sequence, compute:
  sol <- mcLabour(x, te, u)
  H   <- sol[1]
  meMin <- minMonetaryEndowment(sol, x, te)
  HC  <- committedLabour(x, meMin)
  T   <- totalLabour(H, HC)
  # create a dataframe for the current values
  data.frame(
    wage_rate = x, 
    time_endowment = te, 
    utility_level = u, 
    committed_labour = HC,
    uncommitted_labour = H,
    total_labour = T,
    monetary_resource = meMin
    )
}))

# time-compensated monetary endowment
df.tcMe = do.call(rbind, lapply(meSeq, function(x){
  # for each value in the sequence, compute:
  sol <- compensatedLabour(w, x, u)
  H   <- sol[1]
  HC  <- committedLabour(w, x)
  T   <- totalLabour(H, HC)
  teMin <- minTimeEndowment(sol, w, x)
  # create a dataframe for the current values
  data.frame(
    wage_rate = w, 
    monetary_endowment = x, 
    utility_level = u, 
    committed_labour = HC,
    uncommitted_labour = H,
    total_labour = T,
    time_resource = teMin
    )
}))

# time-compensated wage rate
df.tcWr = do.call(rbind, lapply(wSeq, function(x){
  # for each value in the sequence, compute:
  sol <- compensatedLabour(x, me, u)
  H   <- sol[1]
  HC  <- committedLabour(x, me)
  T   <- totalLabour(H, HC)
  teMin <- minTimeEndowment(sol, x, me)
  # create a dataframe for the current values
  data.frame(
    wage_rate = x, 
    monetary_endowment = me, 
    utility_level = u, 
    committed_labour = HC,
    uncommitted_labour = H,
    total_labour = T,
    time_resource = teMin
    )
}))

```

```{r}
plotLabourCurve <- function(df, yvar, ylab, xvar, xlab, fileName) {
  
  # Row 1: labour supply
  df.plot = df %>%
    select(committed_labour, uncommitted_labour, total_labour, all_of(yvar)) %>%
    pivot_longer(cols = c("committed_labour", "uncommitted_labour", "total_labour"), names_to = "type", values_to = "hours") %>%
    mutate(type = factor(type, levels = c("committed_labour", "uncommitted_labour", "total_labour")))
  
  # calculate y min/max dynamically
  # yvals <- df.plot[[yvar]]
  # ymin <- floor(min(yvals, na.rm = TRUE))
  # ymax <- ceiling(max(yvals, na.rm = TRUE))
  
  if (ylab == "Time endowment") {
      ymin <- floor(min(teSeq, na.rm = TRUE))
      ymax <- ceiling(max(teSeq, na.rm = TRUE))
      ybreaks <- pretty(c(ymin, ymax), n = 4)
    } else if (ylab == "Monetary endowment") {
      ymin <- floor(min(meSeq, na.rm = TRUE))
      ymax <- 0
      ybreaks <- pretty(c(ymin, ymax), n = 4)
    } else if (ylab == "Wage rate") {
      ymin <- floor(min(wSeq, na.rm = TRUE)*10)/10
      ymax <- ceiling(max(wSeq, na.rm = TRUE)*10)/10
      ybreaks <- seq(0.4, 1.6, by = 0.4) # pretty(c(ymin, ymax), n = 5)
    } else {
      stop("Error")
    }
  
  # ggplot
  ggp1 = ggplot(df.plot, aes(x = hours, y = .data[[yvar]], color = type)) +
    geom_point() +
    scale_color_manual(
      name = NULL,
      values = c("committed_labour" = "green", 
                 "uncommitted_labour" = "blue", 
                 "total_labour" = "red"),
      labels = c("committed_labour" = "Committed ", 
                 "uncommitted_labour" = "Uncommitted ", 
                 "total_labour" = "Total")
    ) +
    scale_x_continuous(limits = c(0, 12), breaks = seq(0, 12, 4)) +
    scale_y_continuous(limits = c(ymin, ymax), breaks = ybreaks) +
    labs(x = "Labour (time unit)", y = ylab, color = "Type") +
    theme(
      text = element_text(size = 16),
      # legend.position = "inside",
      legend.position = "bottom",
      # legend.position.inside = c(0.85, 0.85), # numeric coordinates
      # legend.background = element_rect(
      #   fill = scales::alpha("white", 0.5),
      #   colour = "black",         # border colour
      #   linewidth = 0.25         # border thickness
      #   ),
      # legend.key = element_rect(fill = NA)
      )
  
  print(ggp1)
  
  # Row 2: Utility, Monetary resource, Time resource
  df.plot = df %>%
    select(all_of(xvar), all_of(yvar))
  
  # calculate min/max dynamically
  xvals <- df.plot[[xvar]]
  if (xlab == "Utility") {
      xmin <- floor(min(xvals, na.rm = TRUE))
      xmax <- ceiling(max(xvals, na.rm = TRUE))
      xbreaks <- pretty(c(xmin, xmax), n = 4)
    } else if (xlab == "Time resource") {
      xmin <- median(xvals) - (max(teSeq) - min(teSeq))/2
      xmax <- median(xvals) + (max(teSeq) - min(teSeq))/2
      xbreaks <- pretty(c(xmin, xmax), n = 4)
    } else if (xlab == "Monetary resource") {
      xmin <- median(xvals) - (max(meSeq) - min(meSeq))/2
      xmax <- median(xvals) + (max(meSeq) - min(meSeq))/2
      xbreaks <- pretty(c(xmin, xmax), n = 4)
    } else {
      # default: dynamic calculation with pretty rounding
      stop("error")
    }

  # ggplot
  ggp2 = ggplot(df.plot, aes(x = .data[[xvar]], y = .data[[yvar]])) +
    geom_point(color = "black")+
    scale_x_continuous(limits = c(xmin, xmax), breaks = xbreaks) +
    scale_y_continuous(limits = c(ymin, ymax), breaks = ybreaks) +
    labs(x = xlab, y = NULL) +
    theme(
      text = element_text(size = 16),
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),  # remove tick labels
      axis.ticks.y = element_blank()  # remove tick marks
      )

  print(ggp2)
  
  # combine plots
  ggp <- plot_grid(
    ggp1,  # left plot
    ggp2,  # right plot
    ncol = 2,         # only 2 columns now
    align = "hv",     # align both horizontally and vertically
    rel_widths = c(1, 1)  # left wider, right narrower
  )
  
  ggsave(filename = fileName, plot = ggp, width = 8, height = 3.5, dpi = 300)
}

# Plot ucTe
plotLabourCurve(
  df.te, # dataframe
  "time_endowment", "Time endowment", # yvar, ylab
  "utility", "Utility", # xvar, xlab
  paste("figure/ch2_ucTeEffect_me", me, "gamma", gamma, ".png",sep = "") # fileName
  )
print("uncompensated time endowment effect")

# Plot ucMe
plotLabourCurve(
  df.me, # dataframe
  "monetary_endowment", "Monetary endowment", # yvar, ylab
  "utility", "Utility", # xvar, xlab
  paste("figure/ch2_ucMeEffect_te", te, "gamma", gamma, ".png",sep = "") # fileName
  )
print("uncompensated monetary endowment effect")

# Plot ucWr
plotLabourCurve(
  df.w, # dataframe
  "wage_rate", "Wage rate", # yvar, ylab
  "utility", "Utility", # xvar, xlab
  paste("figure/ch2_ucWrEffect_me", me, "te", te, "gamma", gamma, ".png",sep = "") # fileName
  )
print("uncompensated wage rate effect")

# Plot mcTe
plotLabourCurve(
  df.mcTe, # dataframe
  "time_endowment", "Time endowment", # yvar, ylab
  "monetary_resource", "Monetary resource", # xvar, xlab
  paste("figure/ch2_mcTeEffect", "utility", u, "gamma", gamma, ".png",sep = "") # fileName
  )
print("monetary-compensated time endowment effect")

# Plot mcWr
plotLabourCurve(
  df.mcWr, # dataframe
  "wage_rate", "Wage rate", # yvar, ylab
  "monetary_resource", "Monetary resource", # xvar, xlab
  paste("figure/ch2_mcWrEffect", "utility", u, "gamma", gamma, ".png",sep = "") # fileName
  )
print("monetary-compensated wage rate effect")

# Plot tcMe
plotLabourCurve(
  df.tcMe, # dataframe
  "monetary_endowment", "Monetary endowment", # yvar, ylab
  "time_resource", "Time resource", # xvar, xlab
  paste("figure/ch2_tcMeEffect", "utility", u, "gamma", gamma, ".png",sep = "") # fileName
  )
print("time-compensated monetary endowment effect")

# Plot tcWr
plotLabourCurve(
  df.tcWr, # dataframe
  "wage_rate", "Wage rate", # yvar, ylab
  "time_resource", "Time resource", # xvar, xlab
  paste("figure/ch2_tcWrEffect", "utility", u, "gamma", gamma, ".png",sep = "") # fileName
  )
print("time-compensated wage rate effect")

```

# ------------------

# Simulate curves: Quadratic

# -----------------

```{r}
# Parameters
alphaC <- -0.15 
betaC  <- 3
alphaL <- -0.01    
betaL  <- 1   
alphaT <- -0.05   
betaT  <- -0.2

# Utility function
utility <- function(H, w, me, te) {
  HC <- committedLabour(w, me)
  C <- consumption(H, w)
  L <- leisure(H, HC, te)
  T <- totalLabour(H, HC)

  if (C < 0 || L <= 0 || T < 0) return(-Inf)  # infeasible solution
  # cat("C, L, T = ", C, L, T, "\n")
  # cat("alphaC * C^2 = ", alphaC * C^2, "\n")
  # cat("betaC * C = ", betaC * C, "\n")
  # cat("alphaL * L^2 = ", alphaL * L^2, "\n")
  # cat("betaL * L = ", betaL * L, "\n")
  # cat("alphaT * T^2 = ", alphaT * T^2, "\n")
  # cat("betaT * T = ", betaT * T, "\n")
  alphaC * C^2 + betaC * C + alphaL * L^2 + betaL * L + alphaT * T^2 + betaT * T
}

# check scales
utility(5, w, me, te)

```

## Utility maximisation problem

We test our utility maximisation problem

```{r}
# ump <- optimize(function(H) utility(H, w, me, te), interval = c(0.01, te - HC - 0.01), maximum = TRUE)
uncompensatedLabour <- function(w, me, te) {
  # Compute committed labour
  HC <- committedLabour(w, me)
  
  # Maximise utility subject to time constraint
  ump <- optimize(
    f = function(H) utility(H, w, me, te),
    interval = c(0.01, te - HC - 0.01),
    maximum = TRUE
  )
  
  # Get optimal H
  H <- ump$maximum
  
  maxU <- ump$objective 
  # print(maxU)
  
  # Return both HC and H
  return(H)
}

# Extract optimal values
HC <- committedLabour(w, me)
ump.H <- uncompensatedLabour(w, me, te)
ump.C <- consumption(ump.H, w)
ump.L <- leisure(ump.H, HC, te)
ump.T <- totalLabour(ump.H, HC)
ump.U <- utility(ump.H, w, me, te)

# Print results
cat("  Time endowment:", te, "\n")
cat("  HC: ", HC, "\n")
cat("  ump.H:", ump.H, "\n")
cat("  ump.U:", ump.U, "\n")
```

## Monetary resource minimisation problem

We test our resource minimisation problem

```{r}
# minMonetaryEndowment function: utilityTarget is only included for the syntax in nloptr. The current calculation doesn't need it.
minMonetaryEndowment <- function(x, w, te, utilityTarget) {
  # x[1] = H, x[2] = L
  if (x[1] <= 0) return(1e6)  # Penalise infeasible
  if (x[2] <= 0) return(1e6)
  w * (x[1] + x[2] - te)
}

# check: it should return me
x = c(ump.H, ump.L)
minMonetaryEndowment(x, w, te)

# Utility function - form 3 - takes L, H and te inputs
mc.utility <- function(x, w, te) {
  
  # x[1] is H, x[2] is L
  
  HC <- te - x[1] - x[2]
  C <- consumption(x[1], w)
  T <- totalLabour(x[1], HC)
  
  # cat("HC: ", HC, "\n")
  # cat("H: ", x[1], "\n")
  
  if (C <= 0 || x[2] <= 0 || T <= 0 || x[1] <= 0) return(-Inf)  # infeasible solution
  # utility(H, w, me, te)
  utility(x[1], w, (-1) * w * HC, te)
}

# check if mc.utility returns ump.U
mc.utility(x, w, te)

# Constraint: utility >= utilityTarget
mc.utilityConstraint <- function(x, w, te, utilityTarget) {
  utilityTarget - mc.utility(x, w, te)
}

# check: it should return 0
mc.utilityConstraint(x, w, te, ump.U)

# Optimise: monetary compensated labour supply
library(nloptr)
mcLabour <- function(w, te, utilityTarget) {
  
  # minimisation
  rmp <- nloptr(
    x0 = c(1,1), # initial guess i.e., starting value
    w = w, te = te,  utilityTarget = utilityTarget, # other inputs
    eval_f = minMonetaryEndowment,
    eval_g_ineq = mc.utilityConstraint,
    lb = c(0.01, 0.01),
    ub = c(Inf, Inf),
    opts = list("algorithm" = "NLOPT_LN_COBYLA", "xtol_rel" = 1e-6)
  )
  
  # Extract optimal values
  rmp.sol <- rmp$solution
  rmp.obj <- rmp$objective
  
  # check objective value
  # cat("Objective value:", rmp.obj, "\n")
  
  # Return compensated H
  return(rmp.sol)
}

rmp.sol = mcLabour(w, te, ump.U)
rmp.H   = rmp.sol[1]
rmp.L   = rmp.sol[2]
meMin = minMonetaryEndowment(rmp.sol, w, te)
rmp.U  = mc.utility(rmp.sol, w, te)

# Results
cat("  Target utility =", ump.U, "\n")
cat("  Achieved utility =", rmp.U, "\n")
cat("  rmp.H:",  rmp.H, "\n")
cat("  rmp.L:",  rmp.L, "\n")
cat("  rmp.me:", meMin, "\n")

```

## Data

```{r}

# simulate wage rates, time endowment, monetary endowment
teSeq <- seq(12, 20, by = 0.5)
meSeq <- seq(-6, -1, by = 0.3)
wSeq  <- seq(0.2, 1.6, by =0.05)

# time endowment
df.te <- do.call(rbind, lapply(teSeq, function(x) {
  H <- uncompensatedLabour(w, me, x)[1]
  HC<- committedLabour(w, me)
  T <- totalLabour(H, HC)
  data.frame(
    wage_rate = w, 
    monetary_endowment = me, 
    time_endowment = x, 
    committed_labour = HC,
    uncommitted_labour = H,
    total_labour = T
    )
}))

# monetary endowment
df.me <- do.call(rbind, lapply(meSeq, function(x) {
  H <- uncompensatedLabour(w, x, te)[1]
  HC<- committedLabour(w, x)
  T <- totalLabour(H, HC)
  data.frame(
    wage_rate = w, 
    monetary_endowment = x, 
    time_endowment = te, 
    committed_labour = HC,
    uncommitted_labour = H,
    total_labour = T
    )
}))

# wage rate
df.w <- do.call(rbind, lapply(wSeq, function(x) {
  H <- uncompensatedLabour(x, me, te)[1]
  HC<- committedLabour(x, me)
  T <- totalLabour(H, HC)
  data.frame(
    wage_rate = x, 
    monetary_endowment = me, 
    time_endowment = te, 
    committed_labour = HC,
    uncommitted_labour = H,
    total_labour = T
    )
}))

```

## Plot

```{r}
plotLabourCurve <- function(df, yvar, ylab, fileName) {
  
  # data
  df.plot = df %>%
    select(committed_labour, uncommitted_labour, total_labour, all_of(yvar)) %>%
    pivot_longer(cols = c("committed_labour", "uncommitted_labour", "total_labour"), names_to = "type", values_to = "hours") %>%
    mutate(type = factor(type, levels = c("committed_labour", "uncommitted_labour", "total_labour")))
  
  # ggplot
  ggp = ggplot(df.plot, aes(x = hours, y = .data[[yvar]], color = type)) +
    geom_point() +
    scale_color_manual(
      name = "Type",
      values = c("committed_labour" = "green", 
                 "uncommitted_labour" = "blue", 
                 "total_labour" = "red"),
      labels = c("committed_labour" = "Committed", 
                 "uncommitted_labour" = "Uncommitted", 
                 "total_labour" = "Total")
    ) +
    scale_x_continuous(limits = c(0, 12), breaks = seq(0, 12, 2)) +
    labs(x = "Labour (time unit)", y = ylab, color = "Type") +
    theme(text = element_text(size = 14))
  
  print(ggp)
  
  ggsave(filename = fileName, plot = ggp, width = 6.3, height = 4.5, dpi = 300)
}

# Plot te vs labour
plotLabourCurve(df.te, "time_endowment", "Time endowment", paste("figure/ch2_quadUtility_teEffect_me", me, "alphaT", alphaT, ".png",sep = ""))

# Plot me vs labour
plotLabourCurve(df.me, "monetary_endowment", "Monetary endowment", paste("figure/ch2_quadUtility_meEffect_me", me, "alphaT", alphaT, ".png",sep = ""))

# Plot wage rate vs labour
plotLabourCurve(df.w, "wage_rate", "Wage rate", paste("figure/ch2_quadUtility_wEffect_me", me, "alphaT", alphaT, ".png",sep = ""))
```

# ------------------

# Policy application

# ------------------

```{r}
# Parameters
alphaC <- -0.15 
alphaL <- -0.01    
betaL  <- 1   
alphaT <- -0.05  
```

## Worker 1

```{r}

# Worker 1
betaC  <- 3
betaT  <- -0.2

# Check optimal values
HC <- committedLabour(w, me)
ump.H <- uncompensatedLabour(w, me, te)
ump.C <- consumption(ump.H, w)
ump.L <- leisure(ump.H, HC, te)
ump.T <- totalLabour(ump.H, HC)
ump.U <- utility(ump.H, w, me, te)

# marginal utilities
mUC = 2 * alphaC * w * ump.H + betaC  
mUCC= 2 * alphaC
D = mUC + w * mUCC * ump.T # segment determinant

# Print results
cat("  --- Worker 1 ---", "\n")
cat("  Time endowment:", te, "\n")
cat("  HC: ", HC, "\n")
cat("  ump.H:", ump.H, "\n")
cat("  ump.T:", ump.T, "\n")
cat("  ump.U:", ump.U, "\n\n")

cat("  Determinant (D)", "\n")
cat("    if D < 0, then occupies the upper segment", "\n")
cat("    otherwise, middle segment", "\n")
cat("  D: ", D, "\n\n")


# Policy: Uncompensated 
me1 = me + 0.1*me 
te1 = te - 0.1*te
headerStr = "  --- After policy (uncompensated) ---"
source("synData9_LSM25_subProcedures01.R")

# Policy: reimbursed commuting costs 
me1 = me + 0*me 
te1 = te - 0.1*te
headerStr = "  --- After policy (reimbursed CC) ---"
source("synData9_LSM25_subProcedures01.R")

# Policy: reimbursed commuting costs + paid commuting time
me1 = me + w*0.1*te 
te1 = te - 0.1*te
headerStr = "  --- After policy (reimbursed CC and paid CT) ---"
source("synData9_LSM25_subProcedures01.R")

# Policy: monetary-compensated
te1 = te - 0.1*te
rmp.sol = mcLabour(w, te1, ump.U)
rmp.H   = rmp.sol[1]
meMin = minMonetaryEndowment(rmp.sol, w, te1)
HC    = committedLabour(w, meMin)
rmp.U = mc.utility(rmp.sol, w, te1)
rmp.T = totalLabour(rmp.H, HC)

cat("  --- After policy (monetary-compensated) ---", "\n")
cat("  Target utility =", ump.U, "\n")
cat("  te: ", te1, "\n")
cat("  meMin:", meMin, "\n")
cat("  meMin (% change):", (meMin - me)/abs(me), "\n")
cat("  meMin - me:", meMin - me, "\n")
cat("  HC:",  HC, "\n")
cat("  rmp.H:",  rmp.H, "\n")
cat("  rmp.T:",  rmp.T, "\n")
cat("  rmp.T (% change):", (rmp.T - ump.T)/ump.T, "\n")
cat("  Achieved utility =", rmp.U, "\n")
cat("  rmp.U (% change):", (rmp.U - ump.U)/ump.U, "\n\n")

# Policy: wage rate-compensated
me1 = me + 0.1*me 
te1 = te - 0.1*te
source("synData9_LSM25_subProcedures02.R")

```

## Worker 2

```{r}
# Worker 2
betaC  <- 2
betaT  <- -0.2

# optimal values
HC <- committedLabour(w, me)
ump.H <- uncompensatedLabour(w, me, te)
ump.C <- consumption(ump.H, w)
ump.L <- leisure(ump.H, HC, te)
ump.T <- totalLabour(ump.H, HC)
ump.U <- utility(ump.H, w, me, te)

# marginal utilities
mUC = 2 * alphaC * w * ump.H + betaC  
mUCC= 2 * alphaC
D = mUC + w * mUCC * ump.T # segment determinant

# Print results
cat("  --- Worker 2 ---", "\n")
cat("  Time endowment:", te, "\n")
cat("  HC: ", HC, "\n")
cat("  ump.H:", ump.H, "\n")
cat("  ump.T:", ump.T, "\n")
cat("  ump.U:", ump.U, "\n\n")

cat("  Determinant (D)", "\n")
cat("    if D < 0, then occupies the upper segment", "\n")
cat("    otherwise, middle segment", "\n")
cat("  D: ", D, "\n\n")

# Policy: Uncompensated 
me1 = me + 0.1*me 
te1 = te - 0.1*te
headerStr = "  --- After policy (uncompensated) ---"
source("synData9_LSM25_subProcedures01.R")

# Policy: reimbursed commuting costs 
me1 = me + 0*me 
te1 = te - 0.1*te
headerStr = "  --- After policy (reimbursed CC) ---"
source("synData9_LSM25_subProcedures01.R")

# Policy: reimbursed commuting costs + paid commuting time
me1 = me + w*0.1*te 
te1 = te - 0.1*te
headerStr = "  --- After policy (reimbursed CC and paid CT) ---"
source("synData9_LSM25_subProcedures01.R")

# Policy: monetary-compensated
te1 = te - 0.1*te
rmp.sol = mcLabour(w, te1, ump.U)
rmp.H   = rmp.sol[1]
meMin = minMonetaryEndowment(rmp.sol, w, te1)
HC    = committedLabour(w, meMin)
rmp.U = mc.utility(rmp.sol, w, te1)
rmp.T = totalLabour(rmp.H, HC)

cat("  --- After policy (monetary-compensated) ---", "\n")
cat("  Target utility =", ump.U, "\n")
cat("  te: ", te1, "\n")
cat("  meMin:", meMin, "\n")
cat("  meMin (% change):", (meMin - me)/abs(me), "\n")
cat("  meMin - me:", meMin - me, "\n")
cat("  HC:",  HC, "\n")
cat("  rmp.H:",  rmp.H, "\n")
cat("  rmp.T:",  rmp.T, "\n")
cat("  rmp.T (% change):", (rmp.T - ump.T)/ump.T, "\n")
cat("  Achieved utility =", rmp.U, "\n")
cat("  rmp.U (% change):", (rmp.U - ump.U)/ump.U, "\n\n")

# Policy: wage rate-compensated
me1 = me + 0.1*me 
te1 = te - 0.1*te
source("synData9_LSM25_subProcedures02.R")

```

## Worker 3

```{r}
# Worker 3
betaC  <- 3
betaT  <- 0.2

# optimal values
HC <- committedLabour(w, me)
ump.H <- uncompensatedLabour(w, me, te)
ump.C <- consumption(ump.H, w)
ump.L <- leisure(ump.H, HC, te)
ump.T <- totalLabour(ump.H, HC)
ump.U <- utility(ump.H, w, me, te)

# marginal utilities
mUC = 2 * alphaC * w * ump.H + betaC  
mUCC= 2 * alphaC
D = mUC + w * mUCC * ump.T # segment determinant

# Print results
cat("  --- Worker 3 ---", "\n")
cat("  Time endowment:", te, "\n")
cat("  HC: ", HC, "\n")
cat("  ump.H:", ump.H, "\n")
cat("  ump.T:", ump.T, "\n")
cat("  ump.U:", ump.U, "\n\n")

cat("  Determinant (D)", "\n")
cat("    if D < 0, then occupies the upper segment", "\n")
cat("    otherwise, middle segment", "\n")
cat("  D: ", D, "\n\n")

# Policy: Uncompensated 
me1 = me + 0.1*me 
te1 = te - 0.1*te
headerStr = "  --- After policy (uncompensated) ---"
source("synData9_LSM25_subProcedures01.R")

# Policy: reimbursed commuting costs 
me1 = me + 0*me 
te1 = te - 0.1*te
headerStr = "  --- After policy (reimbursed CC) ---"
source("synData9_LSM25_subProcedures01.R")

# Policy: reimbursed commuting costs + paid commuting time
me1 = me + w*0.1*te 
te1 = te - 0.1*te
headerStr = "  --- After policy (reimbursed CC and paid CT) ---"
source("synData9_LSM25_subProcedures01.R")

# Policy: monetary-compensated
te1 = te - 0.1*te
rmp.sol = mcLabour(w, te1, ump.U)
rmp.H   = rmp.sol[1]
meMin = minMonetaryEndowment(rmp.sol, w, te1)
HC    = committedLabour(w, meMin)
rmp.U = mc.utility(rmp.sol, w, te1)
rmp.T = totalLabour(rmp.H, HC)

cat("  --- After policy (monetary-compensated) ---", "\n")
cat("  Target utility =", ump.U, "\n")
cat("  te: ", te1, "\n")
cat("  meMin:", meMin, "\n")
cat("  meMin (% change):", (meMin - me)/abs(me), "\n")
cat("  meMin - me:", meMin - me, "\n")
cat("  HC:",  HC, "\n")
cat("  rmp.H:",  rmp.H, "\n")
cat("  rmp.T:",  rmp.T, "\n")
cat("  rmp.T (% change):", (rmp.T - ump.T)/ump.T, "\n")
cat("  Achieved utility =", rmp.U, "\n")
cat("  rmp.U (% change):", (rmp.U - ump.U)/ump.U, "\n\n")

# Policy: wage rate-compensated
me1 = me + 0.1*me 
te1 = te - 0.1*te
source("synData9_LSM25_subProcedures02.R")

```
